preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2021-05-31 BMC_Genome_Biology_submission/grubert_manuscript_figures_tables/manuscript/manuscript_v1.Rmd
More WIP

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@Lupianez:2015aa; @Franke:2016aa; @rao20143d; @Phanstiel:2017uq], Topologically Associating Domains (TADs) [@nora2012spatial; @dixon2012topological; @Hou:2012aa; @sexton2012three], and A/B compartments [@lieberman2009comprehensive; @rao20143d], reviewed in [@Beagan:2020wz]. At the kilobase scale, chromatin loops (corner-dot structures on Hi-C chromatin interaction maps) connect gene promoters with distal enhancers, promoting regulation of gene expression. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. Perhaps the most prominent feature of TADs is that they are demarcated by boundaries, suggesting that most of the mammalian genome is folded into non-overlapping chromatin interaction domains [@nora2012spatial; @dixon2012topological; @Hou:2012aa; @sexton2012three]. Boundaries were found to be enriched in architectural factors such as CTCF (in convergent orientation [@rao20143d; @Tang:2015aa]), RAD21, SMC3 (members of the cohesin complex), YY1, and ZNF143 [@Dixon:2012aa; @Rao:2014aa; @Phillips-Cremins2013:aa; @Corces:2016aa; @Heidari:2014wr]. <!--Emerging evidence has linked TADs and loops to regulation of gene expression, cellular dynamics, differentiation [@phillips2013architectural; @Beagan:2020wz]. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].--> Disruption of boundaries have been reported to promote human cancers [@hnisz2016activation; @taberlay2016three; @Flavahan2016], neurological disorders [@Sun:2018aa], and pathologies of limb development [@Franke:2016aa; @lupianez2016breaking]. Thus, identifying the precise location of boundaries remains a top priority to fully understand the functionality of the human genome.

@@

In contrast to low-resolution Hi-C data, functional/regulatory genome annotations (histone modifications, DNAse I hypersensitive sites, DNA methylation, and transcription factor binding sites (TFBSs)) assembled by the ENCODE [@ENCODE-Project-Consortium:2012aa] and other consortia have been profiled at relatively high resolution (10-300bp). The association of domain boundaries with CTCF and other architectural proteins [@rao20143d; @Tang:2015aa; @Phillips:2009aa] holds promise that these high-resolution marks may inform the more precise location of domain boundaries. Methods like Lollipop used

@@

Random forest has been used in multiple methods (Lollipop, 3dEpiLoop, Peakachu)

@@

Lollipop used motif strength, orientation, loop size, average sequence conservation, in-between features, gene expression.

The ENCODE, NIH Roadmap Epigenomics, FANTOM5, BLUEPRINT, and other members of the International Human Epigenome Consortium (IHEC) [@Stunnenberg:2016aa] have been actively cataloging functional/regulatory genome annotation datasets, such as cell type- and tissue-specific histone modifications, DNAse I hypersensitive sites, DNA methylation, and transcription factor binding sites (TFBSs).

Numerous studies demonstrated the association of domain boundaries with genomic annotations [@Dixon:2015aa; @Whalen:2016aa; @Dekker:2016aa; @Bonev:2016aa; @Wang:2016ab].

@@

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@Lupianez:2015aa; @Franke:2016aa; @rao20143d; @Phanstiel:2017uq], Topologically Associating Domains (TADs) [@nora2012spatial; @dixon2012topological; @Hou:2012aa; @sexton2012three], and A/B compartments [@lieberman2009comprehensive; @rao20143d], reviewed in [@Beagan:2020wz; @Chang:2020vw]. At the kilobase scale, chromatin loops (corner-dot structures on Hi-C chromatin interaction maps) connect gene promoters with distal enhancers, promoting regulation of gene expression. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. Perhaps the most prominent feature of TADs is that they are demarcated by boundaries, suggesting that most of the mammalian genome is folded into non-overlapping chromatin interaction domains [@nora2012spatial; @dixon2012topological; @Hou:2012aa; @sexton2012three]. Boundaries were found to be enriched in architectural factors such as CTCF (in convergent orientation [@rao20143d; @Tang:2015aa]), RAD21, SMC3 (members of the cohesin complex), YY1, and ZNF143 [@Dixon:2012aa; @rao20143d; @Phillips-Cremins2013:aa; @Corces:2016aa; @Heidari:2014wr]. <!--Emerging evidence has linked TADs and loops to regulation of gene expression, cellular dynamics, differentiation [@phillips2013architectural; @Beagan:2020wz]. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].--> Disruption of boundaries have been reported to promote human cancers [@hnisz2016activation; @taberlay2016three; @Flavahan2016], neurological disorders [@Sun:2018aa], and pathologies of limb development [@Franke:2016aa; @lupianez2016breaking]. Thus, identifying the precise location of boundaries remains a top priority to fully understand the functionality of the human genome.

@@

In contrast to low-resolution Hi-C data, functional/regulatory genome annotations (histone modifications, DNAse I hypersensitive sites, DNA methylation, and transcription factor binding sites (TFBSs)) assembled by the ENCODE [@ENCODE-Project-Consortium:2012aa] and other consortia have been profiled at relatively high resolution (10-300bp) [@Dozmorov:2017aa]. The one-dimensional genome annotation data has been used to predict functional chromatin contacts (e.g., HiC-Reg [@Zhang:2019aa]), boundaries of chromatin domains (e.g., Lollipop [@Kai:2018aa], 3DEpiLoop [@Al-Bkhetan:2018aa]), with 48 methods recently reviewed by Tao et al. [@Tao:2021uc]. The association of domain boundaries with CTCF and other genomic annotations [@dixon2012topological; @rao20143d; @Tang:2015aa; @Phillips:2009aa; @phillips2013architectural] holds promise that these high-resolution marks may inform the more precise location of domain boundaries.

3DEpiLoop predicts two types of chromatin loops, CTCF-mediated and RNAPII-mediated. They showed differences in genome annotation signatures with TFBSs best predicting CTCF-mediated loops and histone marks best predicting RNAPII-mediated loops, indicating differences in annotation signatures for different loop types. In their tests, random forest outperformed AdaBoost classification trees, neural networks, SVM, Stochastic Gradient Boosting.

@@

Random forest has been used in multiple methods (Lollipop, 3dEpiLoop, Peakachu, HiC-Reg)

@@

Lollipop used motif strength, orientation, loop size, average sequence conservation, in-between features, gene expression. HiC-Reg uses histone marks, TFBSs, variants of genomic distance (loop size), multi-cell line features.

@@

Boundaries predicted by preciseTAD are dependent on the training data. Numerous methods for boundary detection have been proposed [@Zufferey:2018aa], with a relatively poor replicability among them [@forcato2017comparison]. It is possible that different methods detect functionally distinct boundaries of different chromatin structures [@Beagan:2020wz; @Chang:2020vw]. Although CTCF and architectural proteins are now considered the "canonical" marks of domain boundaries [@rao20143d; @Phillips-Cremins2013:aa; @Corces:2016aa; @Heidari:2014wr], boundaries of TADs discovered by Dixon et al. were also enriched in housekeeping and tRNA genes and active promoters [@Dixon:2012aa]. , 3DEpiLoop predicts two types of chromatin loops when trained on CTCF- and RNAPII ChIA-PET data. These loops showed differences in genome annotation signatures with TFBSs best predicting CTCF-mediated loops and histone marks best predicting RNAPII-mediated loops. Our future goal include predicting boundary types other than those mediated by CTCF and cohesin.

Boundaries have also been defined by different patterns of CTCF orientation [@Nanni:2020aa].
